Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-02-15 ms.Rnw
update

In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
We then show how one can comput the likelihood of a given genealogy.

@@

If $\Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its jump times in the interval $[0,t]$.

@@

Motivating examples:
compartmental models in ID epidemiology,
phylogenetics and systematics.
Wide variety of models are of interest (\cref{fig:examples}).
Linear chain trick.
Migration, superspreading, competition between strains.

@@

In the following, it will be useful to break the jumps of the population process $\Xr_t$ into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
\Cref{fig:markov_state} shows an example with five distinct marks.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

@@

@@

\AAK{[Maybe it would be better to speak instead of $\Hr_t$ being the $\sigma$-algebra generated by the sequences $(\Trh_k,\Xrh_k,\Urh_k)_{k=1}^{\Kr_t}$.
    The only thing we do with $\Hr$ is condition on it.
    One (perhaps minor) problem with this is that we cannot then speak about the measure on histories.
]}

Consider the Markov process $(\Xr_t,\Ur_t)$.
We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{equation*}
  \label{eq:Hdens}
  \begin{aligned}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{aligned}
\end{equation*}
where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.

@@

\begin{figure}
  \begin{center}
    <<geneal0>>=
    freeze(
      seed=522390503,
      simulate(
        "SEIR",
        Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
        S0=10,E0=1,I0=1,R0=0,
        time=10
      )
    ) -> x
    @
    <<geneal1,dependson="geneal0",fig.dim=c(4,2),out.width="50%">>=
    pal <- c("#00274CFF","#FFCB05FF")
    x |> plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
      geom_vline(xintercept=10,linewidth=0.2,color="black")
    @
  \end{center}
  \caption{
    $\time(\G)=10$.
    Two demes, $\Demes=\Set{\func{blue},\func{yellow}}$.
    Extant lineages are denoted with black dots;
    blue dots indicate samples.
    \label{fig:geneal}
  }
\end{figure}

A \emph{genealogy}, $\G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(\G)\in\Rp$, and perhaps a set of samples collected at earlier times (\cref{fig:geneal}A).
A genealogy has a tree- or forest-like structure, with four distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another;
\item \emph{sample nodes}, which represent labeled samples; and
\item \emph{root nodes}, at the base of each tree.
\end{inparaenum}
Each node $a$ is associated with a specific time, $\time(a)$.
In particular, if $a$ is a tip node in $\G$, then $\time(a)=\time(\G)$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $\time(a)\le{\time(a')}$ and $\time(a')-\time(a)$ is the distance between $a$ and $a'$ along the genealogy.
Without loss of generality we assume that $\time(a)=0$ for all root nodes $a$.
We let $\event{\G}{}$ denote the set of all internal and sample node-times of the genealogy $\G$;
we refer to these as \emph{genealogical event times}.

Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme (\cref{fig:geneal}A).

Formally, we define a genealogy, $\G$, to be an ordered pair $(\y,\z)$.
The $\z$ element specifies the genealogy's tree structure, as follows.
Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
Partition \emph{fineness} defines a partial order on $\func{Z}$.
Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
If $\time(\G)=T$, the tree structure of $\G$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
An element $b\in\z_t$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.

The $\y$ element of $\G$ specifies the coloring of branches and location of nodes.
In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
We will also need to count the number of nodes along each lineage:
let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=\time(a)$ to $t=0$.
Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
Formally, we define the coloring component $\y$ as follows.
Let $\Yspace=\Demes\times\Zp$.
For $\eta\in\Yspace$, write $\eta=(\eta^d,\eta^m)$ to denote the two components.
Then, for $\zeta\in\func{Z}$, each $\eta\in\Yspace^{\zeta}$ is a map from $\zeta$ into $\Yspace$.
Now, given a tree $\z:[0,T]\to\func{Z}$ and a map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$,
we say that $\y$ is \emph{compatible} with $\z$ if,
for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$,
$\y$ is \cadlag\ and,
whenever $t_1,t_2\in[0,T]$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$
such that $t_1\le{t_2}$ and $b_1\supseteq{b_2}$,
we have $y^m_{t_1}\le{y^m_{t_2}}$.
Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$ and $\func{Y}_t(\z)\coloneq\CondSet{y_t}{y\in\func{Y}(\z)}$.
\AAK{[The coproduct is a fiber bundle over the space of trees and $\func{Y}(\z)$ is a section of this bundle.]}

@@

\begin{figure}
  <<geneal2,dependson="geneal0",fig.dim=c(4,6),out.width="50%">>=
  pal <- c("#00274CFF","#FFCB05FF")
  plot_grid(
    A=x |>
      plot(
        points=TRUE,prune=FALSE,obscure=FALSE,
        ladderize=FALSE,palette=pal
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    B=x |>
      plot(
        points=TRUE,prune=TRUE,obscure=FALSE,
        ladderize=FALSE,palette=pal
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    C=x |>
      plot(
        points=TRUE,prune=TRUE,obscure=TRUE,
        ladderize=FALSE,palette="#B3B3B3FF"
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    ncol=1,
    align="hv",axis="tblr",
    labels="AUTO"
  )
  @
  \caption{
    Unpruned, pruned, and obscured genealogies from a single realization of the genealogy process induced by the SEIRS model depicted in \cref{fig:example_models,fig:markov_state}.
    \textbf{(A)} A realization of the unpruned genealogy process $\Gr_t$ is shown at $t=10$.
    Tip nodes, corresponding to lineages alive at time $t=10$ are indicated with black points.
    Blue points represent samples;
    green points, internal nodes.
    Branches are colored according to the deme in which the corresponding lineage resided at that point in time:
    blue denotes $\lab{E}$ and yellow, $\lab{I}$.
    \textbf{(B)} The genealogy is \emph{pruned} by deleting all tip nodes and then recursively pruning away childless internal nodes.
    Sample nodes are never removed.
    \textbf{(C)} A genealogy is \emph{obscured} by effacing all deme information from lineage histories:
    the colors are erased, as are extraneous internal nodes.
    See the text (\cref{sec:genealogy,sec:pruning,sec:obscuration}) for more detail.
    \label{fig:upo}
  }
\end{figure}

>>>>

In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
We then show how one can compute the likelihood of a given genealogy.

@@

If $\Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.

@@

Motivating examples:
compartmental models in ID epidemiology,
phylogenetics and systematics.
Wide variety of models are of interest (\cref{fig:example_models}).
Linear chain trick.
Migration, superspreading, competition between strains.

@@

In the following, it will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
\Cref{fig:markov_state} shows an example with five distinct marks.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

@@

\subsubsection{Superspreading model}

\Cref{fig:example_models}D depicts a model of superspreading.
There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}=
    &\frac{\beta(t)\,(I_L+\theta\,I_H)\,(S+1)}{N}\,v(t,S+1,E-1,I_L,I_H,R)
    -\frac{\beta(t)\,(I_L+\theta\,I_H)\,S}{N}\,v(t,S,E,I_L,I_H,R)\\
    &\sigma\,(E+1)\,v(t,S,E+1,I_L-1,I_H,R)
    -\sigma\,E\,v(t,S,E,I_L,I_H,R)\\
    &\eps_{LH}\,(I_L+1)\,v(t,S,E,I_L+1,I_H-1,R)
    -\eps_{LH}\,I_L\,v(t,S,E,I_L,I_H,R)\\
    &\eps_{HL}\,(I_H+1)\,v(t,S,E,I_L-1,I_H+1,R)
    -\eps_{HL}\,I_H\,v(t,S,E,I_L,I_H,R)\\
    &\gamma\,(I_L+1)\,v(t,S,E,I_L+1,I_H,R-1)
    -\gamma\,I_L\,v(t,S,E,I_L,I_H,R)\\
    &\gamma\,(I_H+1)\,v(t,S,E,I_L,I_H+1,R-1)
    -\gamma\,I_H\,v(t,S,E,I_L,I_H,R)\\
    &\omega\,(R+1)\,v(t,S-1,E,I_L,I_H,R+1)
    -\omega\,R\,v(t,S,E,I_L,I_H,R).\\
  \end{split}
\end{equation*}

@@

\AAK{[Maybe it would be better to speak instead of $\Hr_t$ being the random $\sigma$-algebra generated by the sequences $(\Trh_k,\Xrh_k,\Urh_k)_{k=1}^{\Kr_t}$.
    The only thing we do with $\Hr$ is condition on it.
    One (perhaps minor) problem with this is that we cannot then speak about the measure on histories.
]}

Consider the Markov process $(\Xr_t,\Ur_t)$.
We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.

Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{equation*}
  \label{eq:Hdens}
  \begin{aligned}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{aligned}
\end{equation*}
where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.

If $\H$ is such a history, we define $\time(\H)$ to be the right endpoint of its domain and use the notation $\event{\H}=\Set{\Th_1,\dots,\Th_{\K_t}}\subset{[0,\time(\H)]}$ to denote the set of its jump times.

@@

\begin{figure}
  \begin{center}
    <<geneal,fig.dim=c(4,2),out.width="50%">>=
    freeze(
      seed=382490723,
      simulate(
        "SEIR",
        Beta=3,sigma=0.5,gamma=0.2,psi=0.3,omega=0.5,
        S0=15,E0=1,I0=2,R0=0,
        time=10
      )
    ) -> x

    pal <- c("#00274CFF","#FFCB05FF")

    x |> plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
      geom_vline(xintercept=10,linewidth=0.2,color="black")
    @
  \end{center}
  \caption{
    $\time(\G)=10$.
    Two demes, $\Demes=\Set{\func{blue},\func{yellow}}$.
    Tip nodes, denoting extant lineages, are shown in black;
    sample nodes are shown in blue;
    internal nodes are indicated using green dots.
    Note that internal nodes occur both at branch-points and inline.
    Wherever a lineage moves from one deme (color) to another, an internal node occurs, but not conversely.
    \label{fig:geneal}
  }
\end{figure}

A \emph{genealogy}, $\G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(\G)\in\Rp$, and perhaps a set of samples collected at earlier times (\cref{fig:geneal}A).
A genealogy has a tree- or forest-like structure, with four distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another;
\item \emph{sample nodes}, which represent labeled samples; and
\item \emph{root nodes}, at the base of each tree.
\end{inparaenum}
Each node $a$ is associated with a specific time, $\time(a)$.
In particular, if $a$ is a tip node in $\G$, then $\time(a)=\time(\G)$;
if $a$ is a sample node, then $\time(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $\time(a)\le{\time(a')}$ and $\time(a')-\time(a)$ is the distance between $a$ and $a'$ along the genealogy.
Without loss of generality we assume that $\time(a)=0$ for all root nodes $a$.
We let $\event{\G}{}$ denote the set of all internal and sample node-times of the genealogy $\G$;
we refer to these as \emph{genealogical event times}.

Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted with a distinct color for each deme (\cref{fig:geneal}A).
Note that a genealogy will in general have \emph{branch-point nodes}, \ie internal nodes with more than one descendant, but may also have internal nodes with only one descendant.
These occur whenever the color changes along a branch, but can also occur without a color-change.
We refer to such nodes as \emph{inline nodes}.

Formally, we define a genealogy, $\G$, to be an ordered pair $(\y,\z)$.
The $\z$ element specifies the genealogy's tree structure, as follows.
Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
Partition \emph{fineness} defines a partial order on $\func{Z}$.
Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
If $\time(\G)=T$, the tree structure of $\G$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
An element $b\in\z_t$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.
We use the notation $\event{\z}$ to denote the set of all event-times in the tree $\z$.
Note that $\event{\z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
Therefore, $\event{\z}\subseteq{\event{\G}}$.

The $\y$ element of $\G$ specifies the coloring of branches and locations of tip, sample, and internal nodes (including inline nodes).
In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
We will also need to count the number of nodes along each lineage:
let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=\time(a)$ to $t=0$.
Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
Formally, we define the coloring component $\y$ as follows.
Let $\Yspace=\Demes\times\Zp$ and write $\upsilon\in\Yspace$ component-wise as $\upsilon=(\upsilon^d,\upsilon^m)$.
For any $\zeta\in\func{Z}$, an element $\eta\in\Yspace^{\zeta}$ is a map from the partition $\zeta$ into $\Yspace$.
Now, suppose we are given a tree $\z:[0,T]\to\func{Z}$ and a \cadlag\ map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$.
Let $\event{\y}$ denote the set of discontinuities of $\y$.
We say that $\y$ is \emph{compatible} with $\z$ if it is a valid coloring of the tree $\z$.
Formally, compatibility requires the following:
\begin{inparaenum}[(i)]
\item
  $\event{\y}\supseteq\event{\z}$;
\item
  for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$;
\item
  if $0\le{t_1}\le{t_2}\le{T}$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$, and $b_1\supseteq{b_2}$,
  then $y^m_{t_1}\le{y^m_{t_2}}$;
\item $y^m_{t_1}=y^m_{t_2}$ implies $y^d_{t_1}=y^d_{t_2}$.
\end{inparaenum}
Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$ and $\func{Y}_t(\z)\coloneq\CondSet{y_t}{y\in\func{Y}(\z)}$.
\AAK{[These conditions are necessary but not sufficient.]}
\AAK{[The coproduct is a fiber bundle over the $\z$ and $\func{Y}(\z)$ is a section of this bundle.]}

@@

\begin{figure}
  <<upo,fig.dim=c(4,6),out.width="50%">>=
  freeze(
    seed=522390503,
    simulate(
      "SEIR",
      Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
      S0=10,E0=1,I0=1,R0=0,
      time=10
    )
  ) -> x

  pal <- c("#00274CFF","#FFCB05FF")

  plot_grid(
    A=x |>
      plot(
        points=TRUE,prune=FALSE,obscure=FALSE,
        ladderize=FALSE,palette=pal
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    B=x |>
      plot(
        points=TRUE,prune=TRUE,obscure=FALSE,
        ladderize=FALSE,palette=pal
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    C=x |>
      plot(
        points=TRUE,prune=TRUE,obscure=TRUE,
        ladderize=FALSE,palette="#B3B3B3FF"
      )+
      geom_vline(xintercept=10,linewidth=0.2,color="black"),
    ncol=1,
    align="hv",axis="tblr",
    labels="AUTO"
  )
  @
  \caption{
    Unpruned, pruned, and obscured genealogies from a single realization of the genealogy process induced by the SEIRS model depicted in \cref{fig:example_models,fig:markov_state}.
    \textbf{(A)} A realization of the unpruned genealogy process $\Gr_t$ is shown at $t=10$.
    Tip nodes, corresponding to lineages alive at time $t=10$ are indicated with black points.
    Blue points represent samples;
    green points, internal nodes.
    Branches are colored according to the deme in which the corresponding lineage resided at that point in time:
    blue denotes $\lab{E}$ and yellow, $\lab{I}$.
    \textbf{(B)} The genealogy is \emph{pruned} by deleting all tip nodes and then recursively pruning away childless internal nodes.
    Sample nodes are never removed.
    \textbf{(C)} A genealogy is \emph{obscured} by effacing all deme information from lineage histories:
    the colors are erased, as are all inline nodes.
    See the text (\cref{sec:genealogy,sec:pruning,sec:obscuration}) for more detail.
    \label{fig:upo}
  }
\end{figure}
